---
title: "Principle Component Aanalysis - Amanda Oliveira"
output: html_document
---
## Data
The data you will be using comes from the Assistments online intelligent tutoring system (https://www.assistments.org/). It describes students working through online math problems. Each student has the following data associated with them:

- id
- prior_prob_count: How many problems a student has answered in the system prior to this session
- prior_percent_correct: The percentage of problems a student has answered correctly prior to this session
- problems_attempted: The number of problems the student has attempted in the current session
- mean_correct: The average number of correct answers a student made on their first attempt at problems in the current session
- mean_hint: The average number of hints a student asked for in the current session
- mean_attempt: The average number of attempts a student took to answer a problem in the current session
- mean_confidence: The average confidence each student has in their ability to answer the problems in the current session

## Start by uploading the data
```{r}
D1 <- read.csv("Assistments-confidence.csv")

```

## Create a correlation matrix of the relationships between the variables, including correlation coefficients for each pair of variables/features.

```{r}
#You can install the corrplot package to plot some pretty correlation matrices (sometimes called correlograms)

library(ggplot2)
library(GGally)
library(dplyr)

ggpairs(D1, 2:8, progress = FALSE) #ggpairs() draws a correlation plot between all the columns you identify by number (second option, you don't need the first column as it is the student ID) and progress = FALSE stops a progress bar appearing as it renders your plot

ggcorr(D1[,-1], method = c("everything", "pearson")) #ggcorr() doesn't have an explicit option to choose variables so we need to use matrix notation to drop the id variable. We then need to choose a "method" which determines how to treat missing values (here we choose to keep everything, and then which kind of correlation calculation to use, here we are using Pearson correlation, the other options are "kendall" or "spearman")

#Study your correlogram images and save them, you will need them later. Take note of what is strongly related to the outcome variable of interest, mean_correct. 
```

## Create a new data frame with the mean_correct variable removed, we want to keep that variable intact. The other variables will be included in our PCA.

```{r}
D2 <- select(D1, -id, -mean_correct)

```

## Now run the PCA on the new data frame

```{r}
pca <- prcomp(D2, scale. = TRUE)
```

## Although princomp does not generate the eigenvalues directly for us, we can print a list of the standard deviation of the variance accounted for by each component.

```{r}

pca$sdev

#To convert this into variance accounted for we can square it, these numbers are proportional to the eigenvalue

pca$sdev^2

#A summary of our pca will give us the proportion of variance accounted for by each component

summary(pca)

#We can look at this to get an idea of which components we should keep and which we should drop

plot(pca, type = "lines")
```

## Decide which components you would drop and remove them from your data set.

```{r}
# did it in part II!
```


## Part II

```{r}

pca$rotation
biplot(pca)

#Now, create a data frame of the transformed data from your pca.

D3 <- as.data.frame(pca$x)

#Attach the variable "mean_correct" from your original data frame to D3.

D3$mean_correct <- D1$mean_correct

#Now re-run your correlation plots between the transformed data and mean_correct. If you had dropped some components would you have lost important information about mean_correct?

# Although PC6 seems to be the weaker PC among all 6 PCs, it is statistically significant and strongly negatively correlated with mean_correct. Dropping it could make us lose some important information about student performance. 

ggpairs(D3, progress = FALSE)

```
## Now print out the loadings for the components you generated:

```{r}
pca$rotation

#Examine the eigenvectors, notice that they are a little difficult to interpret. It is much easier to make sense of them if we make them proportional within each component

loadings <- abs(pca$rotation) #abs() will make all eigenvectors positive
loadings

#Now examine your components and try to come up with substantive descriptions of what some might represent?

#PC1 seems to account for current interaction with the system (hints and attempts in current session). Students who attempt and ask for hints more frequently also seem to be low in confidence. This measurement of current interaction is moderately influenced by previous interaction. I am inclined to think of this component as a indicator for "over interactive" students. This could be a flag for instructors, as higher scores on PC1 are negatively correlated with mean_correct (see correlation matrix above).

#PC2 seem to be a good indicator of previous performance. Note that it is not highly reliant on any current variables (like current attempts and hint requests). Interestingly, confidence does not seem to be correlated with this measure. I would likely "name" PC2 "Past Performance". Students who scored highly in previous sessions tend to perform better in the current session (see statistically significant correlation in the matrix) 

#PC3 seems to be a good measure of student confidence. Assuming the instructional "main objective" is to maximize current performance, my advice would be not to rely on student confidence as a performance predictor (though I am unsure how confidence was constructed). This PC is not correlated with student performance. 

# Answering to your question from Part1, I would probably use only PC1 and PC2. Although PC6 is indeed statistically significant and strongly negatively correlated with current performance, it does not seem to be any better than using its component variables alone. 

#You can generate a biplot to help you, though these can be a bit confusing. They plot the transformed data by the first two components. Therefore, the axes represent the direction of maximum variance accounted for. Then mapped onto this point cloud are the original directions of the variables, depicted as red arrows. It is supposed to provide a visualization of which variables "go together". Variables that possibly represent the same underlying construct point in the same direction.  

biplot(pca)


```


# Part III  
Also in this repository is a data set collected from TC students (tc-program-combos.csv) that shows how many students thought that a TC program was related to andother TC program. Students were shown three program names at a time and were asked which two of the three were most similar. Use PCA to look for components that represent related programs. Explain why you think there are relationships between these programs.

```{r}

DF1 <- read.csv("tc-program-combos.csv")
DF2 <- select(DF1, -program)

pca2 <- prcomp(DF2, scale. = TRUE)
summary(pca2)

plot(pca2, type = "lines") # from this, it seems reasonable to focus on first 4 PC
pca2$rotation # this is a poor display of results because it omits a lot of rows 

# Inspect loadings 
DF3 <- as.data.frame(pca2$rotation)
DF4 <- select(DF3, 1:4)
program <- rownames(DF4)
DF4 <- cbind(DF4, program)

# PC1: Humanities
PC1 <- select(DF4, PC1, program)
PC1b <- filter(PC1, PC1>=.2)

# PC2: Health
PC2 <- select(DF4, PC2, program)
PC2b <- filter(PC2, PC2>=.2)

# PC3: STEM
PC3 <- select(DF4, PC3, program)
PC3b <- filter(PC3, PC3>=.2)

# PC4 - PC4 seem to be a weaker PC. There are no loadings >=.2
PC4 <- select(DF4, PC4, program)
PC4b <- filter(PC4, PC4>=.1)


# PC1 seems to rely more heavily on programs in Humanities. See PC1b dataframe.
# PC2 groups programs in Healh. See PC2b dataframe.
# PC3 have bigger loadings for STEM programs. See PC3b dataframe.

```





